Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-22 paper/main.tex
minor edits through intro

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high-dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions, interleaved between
watching two course videos from the Khan Academy platform. We apply our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We use
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful, high-resolution
insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

Suppose that a teacher had access to a complete, tangible ``map'' of everything
a student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might it change their
ability to teach that student? Perhaps they might start by checking how well
the student knew the to-be-learned information already, or how much they knew
about related concepts. For some students, they could potentially optimize
their teaching efforts to maximize efficiency by focusing primarily on
not-yet-known content. For other students (or other content areas), it might be
more effective to optimize for direct connections between already known content
and new material. Observing how the student's knowledge changed over time, in
response to their teaching, could also help to guide the teacher towards the
most effective strategy for that individual student.

A common approach to assessing a student's knowledge is to present them with a set of quiz
questions, calculate the proportion they answer correctly, and provide
them with feedback in the form of a simple numeric or letter grade. While such a
grade can provide \textit{some} indication of whether the student has
mastered the to-be-learned material, any univariate measure of performance on a
complex task sacrifices certain relevant information, risks conflating underlying
factors, and so on. For example, consider the relative utility of the
imaginary map described above that characterizes a student's knowledge
in detail, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion-correct scores or letter
grades can instead be used to obtain much more detailed insights into what the
student knows at the time they took the quiz.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience (e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05})
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
clues as to what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models (e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20}) also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, wherein nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analogue of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the specific words contained in those documents have very little
overlap}.

@@

Here we develop a framework for modeling how conceptual knowledge is acquired
during learning. The central idea behind our framework is to use text embedding
models to define the coordinate systems of two maps: a \textit{knowledge
map} that describes the extent to which each concept is currently known, and
a \textit{learning map} that describes changes in knowledge over time. Each
location on these maps represents a single concept, and the maps' geometries
are defined such that related concepts are located nearby in space. We use this
framework to analyze and interpret behavioral data collected from an experiment
that had participants answer sets multiple-choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory
approaches to studying learning and memory (e.g., list-learning studies) often
draw little distinction between memorization and understanding. Instead, these
studies typically focus on whether information is effectively encoded or
retrieved, rather than whether the information is \textit{understood}.
Approaches to studying conceptual learning, such as category learning
experiments, can begin to investigate the distinction between memorization and
understanding, often by training participants to distinguish arbitrary or
random features in otherwise meaningless categorized stimuli. However the
objective of real-world training, or learning from life experiences more
generally, is often to develop new knowledge that may be applied in
\textit{useful} ways in the future. In this sense, the gap between modern
learning theories and modern pedagogical approaches that inform classroom
learning strategies is enormous: most of our theories about \textit{how} people
learn are inspired by experimental paradigms and models that have only
peripheral relevance to the kinds of learning that students and teachers
actually seek~\citep{Macl05, HallGree08}. To help bridge this gap, our study
uses course materials from real online courses to inform, fit, and test models
of real-world conceptual learning. We also provide a demonstration of how our
models can be used to construct ``maps'' of what students know, and how their
knowledge changes with training. In addition to helping to visualize knowledge
(and changes in knowledge), we hope that such maps might lead to real-world
tools for improving how we educate. Taken together, our work shows that
existing course materials and evaluative tools like short multiple-choice
quizzes may be leveraged to gain highly detailed insights into what students
know and how they learn.

>>>>

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high-dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We apply our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We use
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful, high-resolution
insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

Suppose that a teacher had access to a complete, tangible ``map'' of everything
a student knows. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might it change their
ability to teach that student? Perhaps they might start by checking how well
the student knows the to-be-learned information already, or how much they know
about related concepts. For some students, they could potentially optimize
their teaching efforts to maximize efficiency by focusing primarily on
not-yet-known content. For other students (or other content areas), it might be
more effective to optimize for direct connections between already known content
and new material. Observing how the student's knowledge changed over time, in
response to their teaching, could also help to guide the teacher towards the
most effective strategy for that individual student.

A common approach to assessing a student's knowledge is to present them with a set of quiz
questions, calculate the proportion they answer correctly, and provide
them with feedback in the form of a simple numeric or letter grade. While such a
grade can provide \textit{some} indication of whether the student has
mastered the to-be-learned material, any univariate measure of performance on a
complex task sacrifices certain relevant information, risks conflating underlying
factors, and so on. For example, consider the relative utility of the
imaginary map described above that characterizes a student's knowledge
in detail, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion-correct scores or letter
grades can instead be used to obtain far more detailed insights into what a
student knew at the time they took the quiz.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience (e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05}),
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
clues as to what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models (e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20}) also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, wherein nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analogue of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the specific words contained in those documents have very little
overlap}.

@@

Here we develop a framework for modeling how conceptual knowledge is acquired
during learning. The central idea behind our framework is to use text embedding
models to define the coordinate systems of two maps: a \textit{knowledge
map} that describes the extent to which each concept is currently known, and
a \textit{learning map} that describes changes in knowledge over time. Each
location on these maps represents a single concept, and the maps' geometries
are defined such that related concepts are located nearby in space. We use this
framework to analyze and interpret behavioral data collected from an experiment
that had participants answer sets of multiple-choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory
approaches to studying learning and memory (e.g., list-learning studies) often
draw little distinction between memorization and understanding. Instead, these
studies typically focus on whether information is effectively encoded or
retrieved, rather than whether the information is \textit{understood}.
Approaches to studying conceptual learning, such as category learning
experiments, can begin to investigate the distinction between memorization and
understanding, often by training participants to distinguish arbitrary or
random features in otherwise meaningless categorized stimuli. However the
objective of real-world training, or learning from life experiences more
generally, is often to develop new knowledge that may be applied in
\textit{useful} ways in the future. In this sense, the gap between modern
learning theories and modern pedagogical approaches that inform classroom
learning strategies is enormous: most of our theories about \textit{how} people
learn are inspired by experimental paradigms and models that have only
peripheral relevance to the kinds of learning that students and teachers
actually seek~\citep{Macl05, HallGree08}. To help bridge this gap, our study
uses course materials from real online courses to inform, fit, and test models
of real-world conceptual learning. We also provide a demonstration of how our
models can be used to construct ``maps'' of what students know, and how their
knowledge changes with training. In addition to helping to visually capture knowledge
(and changes in knowledge), we hope that such maps might lead to real-world
tools for improving how we educate. Taken together, our work shows that
existing course materials and evaluative tools like short multiple-choice
quizzes may be leveraged to gain highly detailed insights into what students
know and how they learn.
